---
title: "Applied Statistical Programming - Relational Databases"
date: "3/7/2022"
header-includes:
  - \usepackage{amsmath}
  - \usepackage{geometry}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \usepackage{hyperref}
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\textbf{Write R code to answer the following questions. Write the code, and then
    show what the computer returns when that code is run. Thoroughly comment
    your solutions.}

You have until the beginning of class 3/9 at 10:00am to complete the assignment
below. You may use R, but not any online R documentation. Submit the Rmarkdown
and the knitted PDF to Canvas. Have one group member submit the activity with
all group members listed at the top.

\section*{Relational Databases}

Data rarely come in nicely combined CSV files. This exercise gives you practice
combining data sources. You are given three sets of Twitter data that need to be
combined to answer a set of questions below. The data can be found at the
following URLs.

\begin{itemize}
  \item \url{https://github.com/jmontgomery/jmontgomery.github.io/blob/master/PDS/Datasets/Tweets.csv.zip}
  \item \url{https://raw.githubusercontent.com/jmontgomery/jmontgomery.github.io/master/PDS/Datasets/Mayors.csv}
  \item \url{https://raw.githubusercontent.com/jmontgomery/jmontgomery.github.io/master/PDS/Datasets/TwitterMentions.csv}
\end{itemize}

Once you have imported the data, use relational database commands join data as
necessary in order to answer the following questions.
\begin{enumerate}
  \item For each mayor, calculate the number of times they were mentioned
  \item Add to the mentions datset the number of times each mayor tweeted.
  \item Create a combined dataset of all tweets from the tweets and mentions
       data. Subset down to overlapping columns (and rename where needed) to make
       this easy.
  \item Are there any tweets in the mentions dataset from mayors?
\end{enumerate}


**Libraries; Load Data**:
```{r tidy = TRUE}
library(tidyverse)
library(dplyr)

# Load datasets
mayors <- read_csv(file = "https://raw.githubusercontent.com/jmontgomery/jmontgomery.github.io/master/PDS/Datasets/Mayors.csv")
mentions <- read_csv(file = "https://raw.githubusercontent.com/jmontgomery/jmontgomery.github.io/master/PDS/Datasets/TwitterMentions.csv")
tweets <- read_csv("Tweets.csv")
```

**Preliminary Data Analysis**:
```{r tidy = TRUE}
# (i) Rename first column of 'tweets' dataset to 'id':
tweets <- rename(tweets, id = ...1)

# (i) Ensure that each dataset has a matching identifier (or 'key'):
colnames(mayors) #TwitterHandle
colnames(mentions) #MayorHandle
mentions <- rename(mentions, TwitterHandle = MayorHandle)
colnames(tweets) #ScreenName
tweets <- rename(tweets, TwitterHandle = ScreenName)
# FINISHED: all datasets should have the same matching identifier.

# (ii). Ensure that each dataset has a UNIQUE matching identifier:
mayors %>% 
  count(TwitterHandle) %>%
  filter(n > 1)
#   Two duplicates: 'robertgarcialb' and 'rodhiggins2017' have two observations each.
#   Addressed in Drill #3.

mentions %>% 
  count(TweetID) %>%
  filter(n > 1)
#   No duplicates.

tweets %>% 
  count(TweetID) %>%
  filter(n > 1)
#   Two duplicates.
vec <- tweets %>% 
  count(TweetID) %>%
  filter(n > 1) %>%
  pull(TweetID)
vec
format(vec, scientific = FALSE)
rm(vec)
# Two duplicates: 
#   "413450477196816384" for ScreenName 'PATownhall'.
#   "777752770510290944" for ScreenName '	derek_armstead'.

# Investigate tweet "413450477196816384" for ScreenName 'PATownhall'.
tweets %>% 
  mutate(TweetID = format(TweetID, scientific = FALSE)) %>%
  filter(TweetID == "413450477196816384")
# Represents these two tweets:
#   https://twitter.com/PATownhall/status/413450477196816386
#   https://twitter.com/PATownhall/status/413450477196816385
#   Notice that neither tweet has TweetID = 413450477196816384. Correct the TweetIDs:

tweets %>%
  filter(id == 472216) # First tweet
tweets <- tweets %>%
  mutate(TweetID = format(TweetID, scientific = FALSE)) %>%
  mutate(TweetID = replace(TweetID, id == 472216, "413450477196816386")) %>%
  mutate(TweetID = format(TweetID, scientific = TRUE))

tweets %>%
  filter(TweetID == 413450477196816384)
# Second tweet's ID: 472217.
tweets %>%
  filter(id == 472217) # Second tweet
tweets <- tweets %>%
  mutate(TweetID = format(TweetID, scientific = FALSE)) %>%
  mutate(TweetID = replace(TweetID, id == 472216, "413450477196816385")) %>%
  mutate(TweetID = format(TweetID, scientific = TRUE))

tweets %>% 
  count(TweetID) %>%
  filter(n > 1)

# Check last duplicate:
#   "777752770510290944" for ScreenName 'derek_armstead'

tweets %>%
  mutate(count = n()) %>%
  filter(TweetID == 777752770510290944)
#   These tweets do not exist on Twitter. I'm going to remove them.
dim(tweets)
tweets <- tweets %>%
  filter(TweetID != 777752770510290944)
dim(tweets)

tweets %>% 
  count(TweetID) %>%
  filter(n > 1)

# No remaining duplicates in tweets dataset.

```



**# In-class drill 1 \& 2**:
```{r tidy = TRUE}

# Left join 3 variables (FacebookLink, TwitterLink, GenderFemale) into `tweets` dataset by `TwitterHandle` unique identifier:
left <- tweets %>%
  left_join(select(mayors, TwitterHandle, FacebookLink, TwitterLink, GenderFemale),
    by = "TwitterHandle"
  )

# Right join:
right <- tweets %>%
  right_join(select(mayors, TwitterHandle, FacebookLink, TwitterLink, GenderFemale),
    by = "TwitterHandle"
  )

# Inner join (matches only those observations for which the linking variable appears on both tables):
inner <- tweets %>%
  inner_join(select(mayors, TwitterHandle, FacebookLink, TwitterLink, GenderFemale),
    by = "TwitterHandle"
  )

#Full join (keeps all data):
full <- tweets %>%
  full_join(select(mayors, TwitterHandle, FacebookLink, TwitterLink, GenderFemale),
    by = "TwitterHandle"
  )
```

**Drill 3**:
```{r tidy = TRUE}
#  (i). dataframes to only the duplicate mayors:
subsetMayors <- mayors %>%
  filter(TwitterHandle %in% c("robertgarcialb", "rodhiggins2017"))

subsetTweets <- tweets %>%
  filter(TwitterHandle %in% c("robertgarcialb", "rodhiggins2017"))


# (ii). Repeat each join from Drill #2. Explain what happens due to the duplication:

# Left join:
leftsubset <- subsetTweets %>%
  left_join(subsetMayors,
    by = "TwitterHandle"
  )
# EXPLANATION: there are two observations for each tweet. In other words, each of these mayor's tweets are duplicated.

# Right join:
rightsubset <- subsetTweets %>%
  right_join(subsetMayors,
    by = "TwitterHandle"
  )
# EXPLANATION: same as for the left_join of each subset.

# Inner join:
innersubset <- subsetTweets %>%
  inner_join(subsetMayors,
    by = "TwitterHandle"
  )
innersubset <- arrange(innersubset, TweetID)
# EXPLANATION: same as for left_join and right_join above.

#Full join:
fullsubset <- subsetTweets %>%
  full_join(subsetMayors,
    by = "TwitterHandle"
  )
# EXPLANATION: same as for the three joins above.

# OVERALL EXPLANATION: all joins result in duplicate copies of each tweet by these mayors. This is because we lack a unique matching 'key' for these mayors for which we can match across datasets.
```

**In-Class Assignment**:
```{r tidy = TRUE}
# Item 1 -----------------------------------------------------------------------
## Create count for times mayors were mentioned
mentions_count <- mentions %>%
  group_by(TwitterHandle) %>%
  mutate(TimesMentioned = n()) %>%
  select(TwitterHandle, TimesMentioned) %>%
  unique()

mayors_mentions_count <- mayors %>%
  left_join(select(mentions_count, TwitterHandle, TimesMentioned),
    by = "TwitterHandle"
  )

# Item 2 -----------------------------------------------------------------------
mayors_tweets <- tweets %>%
  group_by(TwitterHandle) %>%
  mutate(times_tweeted = n()) %>%
  select(TwitterHandle, times_tweeted) %>%
  unique()

mentions <- mentions %>% left_join(mayors_tweets, by = "TwitterHandle")

# Item 3 -----------------------------------------------------------------------
## Columns to remove:
## - ScreenName
## - Column 1 from each one?

tweets_subset <- tweets %>%
  select(-1)

mentions_subset <- mentions %>%
  select(-c(1, 3, 17))

## DON'T RUN. It's currently not working.
# tweets_mentions <- tweets_subset %>%
#   left_join(mentions_subset, by = "TwitterHandle")

# Item 4------------------------------------------------------------------------
## Is this an inner join?
inner_tweets <- tweets_subset %>%
  inner_join(mentions_subset)

nrow(inner_tweets)

```
